Bayesian Exploration in Deep Reinforcement Learning

### 2022-08-29 papers/CDRL/sample_paper.tex
TEMPORARY DEPLOY COMMIT

Unfortunately, this is still a hard optimization problem. We can, however,
manipulate the KL-term to make the optimization problem tractable.
\begin{align*}
  &\KL{q(\vtheta)}{p(\vtheta \vert \data)} =
  \E{q(\vtheta)}{\log \frac{q(\vtheta)}{p(\vtheta \vert \data)}}\\
  &= \E{q(\vtheta)}{\log q(\vtheta)} - \E{q(\vtheta)}{\log p(\vtheta \vert \data)}\\
  &= \E{q(\vtheta)}{\log q(\vtheta)} - \E{q(\vtheta)}{\log
  \frac{p(\vtheta, \data)}{p(\data)}}\\
  &= \E{q(\vtheta)}{\log q(\vtheta)} - \E{q(\vtheta)}{\log p(\vtheta, \data)} +
  \E{q(\vtheta)}{\log p(\data)}\\
  &= \E{q(\vtheta)}{\log q(\vtheta)} - \E{q(\vtheta)}{\log p(\vtheta, \data)} +
  \log p(\data)\\
\end{align*}
This shows that minimizing the KL-divergence is the same as maximizing the
evidence lower bound (\textbf{ELBO}),
\begin{equation*}
  \mathrm{ELBO}(q; \vtheta, \data) = \E{q(\vtheta)}{\log p(\vtheta, \data)} -
  \E{q(\vtheta)}{\log q(\vtheta)}.
\end{equation*}

@@

Let \(\rvZ\) be a random variable defined over the support \(S \subseteq \R^n\)
and a continous probability density function \(q(\rvz)\). Further, let \(v\) be
a continous, differentiable and invertible function from \(S\) to
\(T \in \R^n\), if we let \(\rvZ^\prime = v(\rvZ)\), then the probability
function \(q(\rvz^\prime)\) of \(\rvZ^\prime\) is given by
\begin{align}
  q(\rvz^\prime) &= q(\rvz) \left\lvert \det \left( \frac{\partial \rvz}{
    \partial \rvz^\prime} \right) \right\rvert \nonumber\\
  &= q(v^{-1}(\rvz^\prime)) \left\lvert \det \left( \frac{\partial}{\partial
  \rvz^\prime} v^{-1}(\rvz^\prime) \right) \right\rvert
  \label{eq:change_var_form1}
  % &= q(v^{-1}(\rvz^\prime)) \left\lvert \det \left( \frac{\partial}{\partial
  % \rvz} v(\rvz) \right) \right\rvert^{-1}
  % \label{eq:change_var_form2}
\end{align}

@@

\begin{align}
  q(\rvz^\prime \vert \rvx) &= q(\rvz \vert \rvx) \left\lvert \det
  \left( \frac{\partial \rvz}{\partial \rvz^\prime} \right) \right\rvert
  \nonumber\\
  &= q(v^{-1}(\rvz^\prime, \rvx)) \left\lvert \det \left( \frac{\partial}
  {\partial \rvz^\prime} v^{-1}(\rvz^\prime, \rvx) \right) \right\rvert.
  \label{eq:cond_change_var_form1}
  % &= q(v^{-1}(\rvz^\prime; \rvx)) \left\lvert \det \left( \frac{\partial}
  % {\partial \rvz} v(\rvz; \rvx) \right) \right\rvert^{-1}
\end{align}

@@

\begin{algorithm}
  \caption{Flow-RL Loss}\label{alg:LTM}
  \textbf{Require:} \(N, v, v^{-1}, \theta, \tilde{\theta}\)\\
  \textbf{input} \(\rvx, a, r, \rvx^\prime, \gamma\)\\
  \hspace*{1em} {\color{gray} \# Take \(N\) samples from the value distribution.}\\
  \hspace*{1em} \textbf{for} \(i = 1 \dots N\)\\
  \hspace*{2em} \(\epsilon_i \sim \N \left(0, I_{\lvert \A \rvert} \right)\)\\
  \hspace*{2em} \textbf{for} \(a^\prime \in \A\)\\
  \hspace*{3em} \(Z_{\tilde{\theta}}(\rvx^\prime, a^\prime)_i \gets v_{\tilde{\theta}}(\epsilon_i; \rvx^\prime)_{a^\prime}\)\\
  \hspace*{2em} \textbf{end}\\
  \hspace*{1em} \textbf{end}\\
  \\\hspace*{1em} {\color{gray} \# Compute greedy next action.}\\
  \hspace*{1em} \(a^* \gets \argmax_{a^\prime} \frac{1}{N} \sum_{i=1}^N Z_{\tilde{\theta}}(\rvx^\prime, a^\prime)_i\)\\
  \\\hspace*{1em} {\color{gray} \# Calculate loss terms for each sample.}\\
  \hspace*{1em} \textbf{for} \(i = 1 \dots N\)\\
  \hspace*{2em} {\color{gray} \# Apply the distributional Bellman operator.}\\
  \hspace*{2em} \(\hat{Z}_{\tilde{\theta}}(\rvx, a)_i \gets r + \gamma Z_{\tilde{\theta}}(\rvx^\prime, a^*)_i\)\\
  \hspace*{2em} {\color{gray} \# Calculate log-likelihood under \(\N(0,1)\).}\\
  \hspace*{2em} \(\delta_i \gets \frac{1}{2} v_{\theta}^{-1} 
  \left(\hat{Z}_{\tilde{\theta}}(\rvx, a)_i; \rvx \right)^2\)\\
  \hspace*{2em} {\color{gray} \# Calculate log-derivative of \(v^{-1}_\theta\).}\\
  \hspace*{2em} \(\rho_i \gets \log \left\lvert \frac{\partial v_{\theta}^{-1} \left(\hat{Z}_{\tilde{\theta}}(\rvx, a)_i; \rvx \right)}{\hat{Z}_{\tilde{\theta}}(\rvx, a)_i} \right\rvert\)\\
  \hspace*{1em} \textbf{end}\\
  \textbf{output} \(\frac{1}{N}\sum_{i=1}^N \delta_i - \rho_i\)
\end{algorithm}

@@

Equation~\ref{eq:luddeflow} shows the definitiona of the forward transformation
\(v_{w, b, c, d}\), with 4 free parameters \(w, b, c, d\). Unlike the planar
flow, this function is invertible and differentiable given any values for these
parameters. Equation~\ref{eq:luddeflow_inv} and
Equation~\ref{eq:luddeflow_grad} show the closed form gradient.

% \begin{equation}
%   Z_\theta(\rvx,a) = v_\theta^{-1}(z, \rvx^\prime)_a
% \end{equation}
% \begin{equation*}
%   z \sim \N(0, I); \quad \rvx^\prime = h(\rvx)
% \end{equation*}

% \begin{equation*}
%   s(x) = \log(1 + \exp(x))
% \end{equation*}
% \begin{equation}
%   v_{w, b, c, d}(z) = \sign(z + c) \left[ s(\lvert w(z + c) \rvert + b) 
%   - s(b) \right] + d \label{eq:luddeflow}
% \end{equation}

% \begin{equation*}
%   s^{-1}(x) = \log(\exp(x) - 1); \quad x > 0
% \end{equation*}
% \begin{equation}
%   v^{-1}_{w, b, c, d}(z^\prime) = \frac{\sign(z^\prime - d)}{w}
%   [ s^{-1}(\lvert z^\prime - d \rvert + s(b)) - b] - c \label{eq:luddeflow_inv}
% \end{equation}

% \begin{equation}
%   \frac{\partial v^{-1}_{w,b,c,d}}{\partial z^\prime} = \frac{(e^b + 1) 
%   e^{\lvert z^\prime - d \rvert}}{w (e^b + 1)e^{\lvert z^\prime - d \rvert} - w}
%   \label{eq:luddeflow_inv_grad}
% \end{equation}

@@

Figure~\ref{fig:luddeflow} shows the effect of the free parameters. A simple
description is \(w\) changes the slope of the function, \(b\) changes the
\emph{linearity} of the function (as \(b\) goes to infinity, \(v\) approaches a
linear function), \(c\) shifts the horizontally, and \(d\) shifts the function
vertically.

\begin{figure}[ht]
  \vspace{.3in}
  % \centerline{\fbox{This figure intentionally left non-blank}}
  \includegraphics[width=\columnwidth]{luddeflow.png}
  \vspace{.3in}
  \caption{Illustration of the effect the free parameters have on the function).
  }\label{fig:luddeflow}
\end{figure}

>>>>

Unfortunately, this is still a hard optimization problem. We can, however,
manipulate the KL-term to make the optimization problem tractable.
\begin{align*}
   & \KL{q(\vtheta)}{p(\vtheta \vert \data)} =
  \E{q(\vtheta)}{\log \frac{q(\vtheta)}{p(\vtheta \vert \data)}}                     \\
   & = \E{q(\vtheta)}{\log q(\vtheta)} - \E{q(\vtheta)}{\log p(\vtheta \vert \data)} \\
   & = \E{q(\vtheta)}{\log q(\vtheta)} - \E{q(\vtheta)}{\log
  \frac{p(\vtheta, \data)}{p(\data)}}                                                \\
   & = \E{q(\vtheta)}{\log q(\vtheta)} - \E{q(\vtheta)}{\log p(\vtheta, \data)} +
  \E{q(\vtheta)}{\log p(\data)}                                                      \\
   & = \E{q(\vtheta)}{\log q(\vtheta)} - \E{q(\vtheta)}{\log p(\vtheta, \data)} +
  \log p(\data)                                                                      \\
\end{align*}
This shows that minimizing the KL-divergence is the same as maximizing the
evidence lower bound (\textbf{ELBO}),
\begin{equation*}
  \mathrm{ELBO}(q; \vtheta, \data) = \E{q(\vtheta)}{\log p(\vtheta, \data)} -
  \E{q(\vtheta)}{\log q(\vtheta)}.
\end{equation*}

@@

Let \(\rvZ\) be a random variable defined over the support \(S \subseteq \R^n\)
and a continous probability density function \(q(\rvz)\). Further, let \(v\) be
a continous, differentiable and invertible function from \(S\) to
\(T \in \R^n\), if we let \(\rvZ^\prime = v(\rvZ)\), then the probability
function \(q(\rvz^\prime)\) of \(\rvZ^\prime\) is given by
\begin{align}
  q(\rvz^\prime) & = q(\rvz) \left\lvert \det \left( \frac{\partial \rvz}{
  \partial \rvz^\prime} \right) \right\rvert \nonumber                                       \\
                 & = q(v^{-1}(\rvz^\prime)) \left\lvert \det \left( \frac{\partial}{\partial
    \rvz^\prime} v^{-1}(\rvz^\prime) \right) \right\rvert
  \label{eq:change_var_form1}
  % &= q(v^{-1}(\rvz^\prime)) \left\lvert \det \left( \frac{\partial}{\partial
  % \rvz} v(\rvz) \right) \right\rvert^{-1}
  % \label{eq:change_var_form2}
\end{align}

@@

\begin{align}
  q(\rvz^\prime \vert \rvx) & = q(\rvz \vert \rvx) \left\lvert \det
  \left( \frac{\partial \rvz}{\partial \rvz^\prime} \right) \right\rvert
  \nonumber                                                                                          \\
                            & = q(v^{-1}(\rvz^\prime, \rvx)) \left\lvert \det \left( \frac{\partial}
  {\partial \rvz^\prime} v^{-1}(\rvz^\prime, \rvx) \right) \right\rvert.
  \label{eq:cond_change_var_form1}
  % &= q(v^{-1}(\rvz^\prime; \rvx)) \left\lvert \det \left( \frac{\partial}
  % {\partial \rvz} v(\rvz; \rvx) \right) \right\rvert^{-1}
\end{align}

@@

\begin{algorithm}
  \caption{Flow-RL Loss}\label{alg:LTM}
  \textbf{Require:} \(N, v, v^{-1}, \theta, \tilde{\theta}\)\\
  \textbf{input} \(\rvx, a, r, \rvx^\prime, \gamma\)\\
  \hspace*{1em} {\color{gray} \# Take \(N\) samples from the value distribution.}\\
  \hspace*{1em} \textbf{for} \(i = 1 \dots N\)\\
  \hspace*{2em} \(\epsilon_i \sim \N \left(0, 1 \right)\)\\
  \hspace*{2em} \textbf{for} \(a^\prime \in \A\)\\
  \hspace*{3em} \(Z_{\tilde{\theta}}(\rvx^\prime, a^\prime)_i \gets v_{\tilde{\theta}}(\epsilon_i; \rvx^\prime)_{a^\prime}\)\\
  \hspace*{2em} \textbf{end}\\
  \hspace*{1em} \textbf{end}\\
  \\\hspace*{1em} {\color{gray} \# Compute greedy next action.}\\
  \hspace*{1em} \(a^* \gets \argmax_{a^\prime} \frac{1}{N} \sum_{i=1}^N Z_{\tilde{\theta}}(\rvx^\prime, a^\prime)_i\)\\
  \\\hspace*{1em} {\color{gray} \# Calculate loss terms for each sample.}\\
  \hspace*{1em} \textbf{for} \(i = 1 \dots N\)\\
  \hspace*{2em} {\color{gray} \# Apply the distributional Bellman operator.}\\
  \hspace*{2em} \(\hat{Z}_{\tilde{\theta}}(\rvx, a)_i \gets r + \gamma Z_{\tilde{\theta}}(\rvx^\prime, a^*)_i\)\\
  \hspace*{2em} {\color{gray} \# Calculate negative log-likelihood loss term.}\\
  \hspace*{2em} \(\delta_i \gets \frac{1}{2} v_{\theta}^{-1}
  \left(\hat{Z}_{\tilde{\theta}}(\rvx, a)_i; \rvx \right)^2\)\\
  \hspace*{2em} {\color{gray} \# Calculate log-derivative of \(v^{-1}_\theta\).}\\
  \hspace*{2em} \(\rho_i \gets \log \left\lvert \frac{\partial v_{\theta}^{-1} \left(\hat{Z}_{\tilde{\theta}}(\rvx, a)_i; \rvx \right)}{\hat{Z}_{\tilde{\theta}}(\rvx, a)_i} \right\rvert\)\\
  \hspace*{1em} \textbf{end}\\
  \textbf{output} \(\frac{1}{N}\sum_{i=1}^N \delta_i - \rho_i\)
\end{algorithm}

@@

Equation~\ref{eq:luddeflow} shows the definitiona of the forward transformation
\(v(\cdot; b, c, d)\), with 3 free parameters \(b, c, d\). Unlike the planar
flow, this function has a closed form inverse and is differentiable given any values for these
parameters, hence no reparameterization is necessary. Equation~\ref{eq:luddeflow_inv} and
Equation~\ref{eq:luddeflow_grad} show the closed form inverse and gradient respectively.

@@

Figure~\ref{fig:luddeflow} shows the effect of the free parameters. A simple
description is that \(b\) changes the
\emph{linearity} of the function (\(v\) is linear when \(b = 0\), an expansive
mapping when \(b < 0\) and a contraction when \(b > 0\)), \(c\) shifts \(v\) horizontally, and \(d\) shifts \(v\) vertically.

\begin{figure}[ht]
  \vspace{.3in}
  % \centerline{\fbox{This figure intentionally left non-blank}}
  \includegraphics[width=\columnwidth]{new_flow.pdf}
  \vspace{.3in}
  \caption{Illustration of the effect the free parameters have on the function.
  }\label{fig:luddeflow}
\end{figure}
